Concurrent Programming in Harmony

### 2021-01-31 book/paper.tex
Pass over Ch 5

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
thread also tries to access the binary tree.
Typically, you want some invariant property of the data structure to hold
at the beginning and at the end of the critical section, but in the middle
the invariant may be temporarily broken---this is no issue as critical
sections guarantee that no other
thread will be able to see it.
An implementation of a data structure that can be safely accessed by multiple
threads (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}%
\index{thread}%

@@

Unfortunately, if we run the program through Harmony,
we find that the assertion still fails.
\autoref{fig:naiveLockhtml} shows the Harmony output.
Thread~0 finds that the lock is available, but just before it stores
\texttt{True} in \textit{lockTaken} in instruction~9, thread~1 gets
to run.
(Recall that you can hover your mouse over a machine instruction in order
to see what it does.)
Because \textit{lockTaken} is still \texttt{False}, it too
believes it can acquire the lock, and stores \texttt{True} in
\texttt{lockTaken} and moves on to the critical section.
Finally, thread~0 moves on, also stores \texttt{True} and also moves
into the critical section.  Thread~0 is the one that detects the
problem.
The \textit{lockTaken} variable
suffers from the same sort of race condition as the \textit{count} variable
in \autoref{fig:inc}: testing and setting the lock
consists of several instructions.  It is thus possible
for both threads to believe the lock is available and to obtain the lock
at the same time.

@@

However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
it turns out the solution
does have a problem: if both try to enter the critical section at the same
time, they may end up waiting for one another indefinitely.  Thus the
solution violates \emph{progress}.

@@

Run the program through Harmony.  It turns out that this solution also violates
\emph{progress}, albeit for a different reason:
if thread~$i$ terminates instead of entering the critical section,
thread $1-i$, politely, ends up waiting indefinitely for its turn.
Too bad, it would have been a great solution if both threads try to enter the
critical section continually.

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert
\begin{code}
\texttt{await False}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure out how to develop a solution that satisfies both
mutual exclusion and progress---as
we will find out, it is possible but not easy.
\end{problems}

>>>>

Critical sections are useful when accessing a shared data
structure, particularly when that access requires multiple underlying
machine instructions.  A counter is a very simple example of
a data structure, but as we have seen it too requires multiple instructions.
A more involved one would be accessing a binary tree.
Adding a node to a binary tree, or re-balancing a tree, often requires
multiple operations.  Maintaining ``consistency'' is certainly much easier
(although not necessarily impossible) if during this time no other
thread also tries to access the binary tree.
Typically, you want some invariant property of the data structure to hold
at the beginning and at the end of the critical section, but in the middle
the invariant may be temporarily broken---this is not a problem as critical
sections guarantee that no other
thread will be able to see it.
An implementation of a data structure that can be safely accessed by multiple
threads (or threads) and free of race conditions is called \emph{thread-safe}.
\index{thread safety}%
\index{thread}%

@@

Unfortunately, if we run the program through Harmony,
we find that the assertion still fails.
\autoref{fig:naiveLockhtml} shows the Harmony output.
Thread~0 finds that the lock is available, but just before it stores
\texttt{True} in \textit{lockTaken} in instruction~9, thread~1 gets
to run.
(Recall that you can hover your mouse over a machine instruction in order
to see what it does.)
Because \textit{lockTaken} is still \texttt{False}, it too
believes it can acquire the lock, and stores \texttt{True} in
\textit{lockTaken} and moves on to the critical section.
Finally, thread~0 moves on, also stores \texttt{True}
into \textit{lockTaken} and also moves
into the critical section.  Thread~0 is the one that detects the
problem.
The \textit{lockTaken} variable
suffers from the same sort of race condition as the \textit{count} variable
in \autoref{fig:inc}: testing and setting the lock
consists of several instructions.  It is thus possible
for both threads to believe the lock is available and to obtain the lock
at the same time.

@@

However, if you run the program through Harmony (\autoref{fig:naiveflagshtml}),
it turns out the solution
does have a problem: if both try to enter the critical section at the same
time, they may end up waiting for one another indefinitely.
(This is a form of \emph{deadlock}, which will be discussed in
\autoref{ch:deadlock}.)
Thus the
solution violates \emph{progress}.

@@

Run the program through Harmony.  It turns out that this solution also violates
\emph{progress}, albeit for a different reason:
if thread~$i$ terminates instead of entering the critical section,
thread $1-i$, politely, ends up waiting indefinitely for its turn.
Too bad, because it would have been a great solution if both
threads try to enter the critical section ad infinitum.

\section*{Exercises}
\begin{problems}
\item Run \autoref{fig:cs} using Harmony.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now insert
\begin{code}
\texttt{await False}
\end{code}
just before entering the critical section
in \autoref{fig:cs} and run Harmony again.
Mutual exclusion is guaranteed but progress is violated.
Harmony should print a trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with Harmony and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure out how to develop a solution that satisfies both
mutual exclusion and progress---as
we will find out, it is possible but not obvious.
\end{problems}
